---
title: "Automation"
author: ''
date: ''
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>", highlight = TRUE, error = TRUE,
                      fig.align = "center")
```

## Main ideas

- Use `for` to iterate code

- Use `map_*()` functions to iterate across data frames

- Iteration best practices

## Packages

```{r}
library(tidyverse)
library(broom)
```

# Notes

## `for` loops

- One tool for reducing duplication is functions. Another tool is iteration
  via `for` loops.
  
- This will help if you need to do the same thing to multiple inputs.

- For example, we can iterate through elements of a vector and evaluate code
  based on each vector element's value.
  
### Motivating example

Let's create a small tibble `x`.

```{r}
x <- tibble(
  col_a = c(3, -1, 0, 10),
  col_b = c(2, -2, 2, -2),
  col_c = c(8, sqrt(131), log(4), 33),
  col_d = 1:4
)
```

Suppose you want to compute the mean of each column in `x`.

A first attempt might be as follows.

```{r}
mean_col_a <- x %>% 
  pull(1) %>% 
  mean()

mean_col_b <- x %>% 
  pull(2) %>% 
  mean()

mean_col_c <- x %>% 
  pull(3) %>% 
  mean()

mean_col_d <- x %>% 
  pull(4) %>% 
  mean()

c(mean_col_a, mean_col_b, mean_col_c, mean_col_d)
```

```{r}
x %>% 
  summarise(
    mean_col_a = mean(col_a),
    mean_col_b = mean(col_b),
    mean_col_c = mean(col_c),
    mean_col_d = mean(col_d)
  )
```

Or a non-tidyverse way may be as follows.

```{r}
mean(x$col_a)
mean(x$col_b)
mean(x$col_c)
mean(x$col_d)
```

This is still a lot of copied code. Imagine if we had a tibble with 1,000
columns.

How can we automate this process?

### `for`-loop construction and syntax

Looking at our previous code, we see that the only variation is with regards
to the column index being pulled. A `for` loop can easily automate our process
from the previous slide.

**First, create an output object.** This is where we will save our results.

**Second, define the loop sequence.** Here `i` is a looping variable, in each
run of the loop `i` will be assigned a different value in the vector
`c(1, 2, 3, 4)`.

**Third, add the loop's body.** This is the code that does the work.

```{r}
# allocate output object, call it results
results <- numeric(4)

# define loop and looping sequence
for (i in c(1, 2, 3, 4)) {
  # print(i)
  cat("The value of i is", i, "\n")
  results[i] <- x %>% 
    pull(i) %>% 
    mean()
  cat("The mean is", results[i], "\n")
  Sys.sleep(3)
}
```

```{r}
results <- numeric(length(x))
for (i in 1:length(x)) {
  results[i] <- x %>% 
    pull(i) %>% 
    mean()
}
```

This is a small example, but it's easy to see the benefits of using a `for`
loop if we needed to scale this computation to 100 columns. All we would
have to change is our loop sequence.

### Example

Let's create a function called `length_unique()`. It will determine the
number of unique values in a vector.

```{r}
length_unique <- function(x) {
  length(unique(x))
}
```

```{r}
a <- c(1, 1, 1, 1, 5, 2, 2, 2, 0)
length(a)
length_unique(a)
```

Write a loop that outputs the number of unique values in each column in a
data frame. Test your loop out on the `storms` data frame in `dplyr`.

```{r}
output <- numeric(length(storms))

for (j in seq(length(storms))) {
  output[j] <- storms %>% 
    pull(j) %>% 
    length_unique()
}

output
```

### `for`-loop best practices

1. Always initialize your result vector with as many elements as you will have
   results before you begin your loop.
   
2. Avoid using loops when vectorization works. For example, in R we do not
   need a loop to sum all the elements of `c(4, 1, -3, 0, 44, 9)`, just use
   `sum()`.
   
A failure to allocate result object

```{r}
results <- NULL
x <- rnorm(n = 100000)

system.time({
for (i in 1:length(x)) {
  results <- c(results, x[i] ^ 2)
}
})
```

```{r}
results <- numeric(100000)
x <- rnorm(n = 100000)

system.time({
for (i in 1:length(x)) {
  results[i] <- x[i] ^ 2
}
})
```

## `purrr` and `map_*()`

The `purrr` package has functions that facilitate automation and mostly
eliminate your need to write `for`-loops.

Recall our small data frame `x`.

```{r}
x <- tibble(
  col_a = c(3, -1, 0, 10),
  col_b = c(2, -2, 2, -2),
  col_c = c(8, sqrt(131), log(4), 33),
  col_d = 1:4
)
```

Instead of a `for`-loop to compute the mean of each column, we could use
`map_dbl()`.

```{r}
map_dbl(x, mean)
```

What is this doing?

```{r}
c(mean(x$col_a), mean(x$col_b), mean(x$col_c), mean(x$col_d))
```

Function `map_dbl()` expects each function call to result in a single
value (vector of length 1). Depending on what you expect your output to be
you can use some similar variants:

- `map_lgl()` for logical results
- `map_chr()` for character results
- `map_dbl()` for double results
- `map_int()` for integer results

### Example

Suppose you want to apply a function to each column of your data frame where
it only modifies the data frame. For example, suppose you want to standardize
your variables before fitting a linear model. Function `modify()` from
`purrr` can help you do just that.

Fit a linear model with `mpg` as the response and `hp` and `wt` as
predictors from `mtcars`. Standardize all variables.

```{r}
standardize <- function(x) {
  (x - mean(x)) / sd(x)
}
```

```{r}
mtcars %>% 
  select(mpg, hp, wt) %>% 
  modify(standardize) %>% 
  lm(mpg ~ -1 + hp + wt, data = .) %>% 
  tidy()
```

## Practice

(1) Write a `for`-loop to check the type of each variable in `gapminder` from
    the `gapminder` package. Function `typeof()` can be used to check the type.
    
```{r}
library(gapminder)

types <- character(length = length(gapminder))

for (i in 1:length(gapminder)) {
  types[i] <- gapminder %>% pull(i) %>% typeof()
}
```

(2) Redo the previous exercise, but this time use a `map_*()` variant.

```{r}
map_chr(gapminder, typeof)
```

(3) Write a function that returns the number of `NA` values in a vector. Use
    a `purrr` function to then check how many `NA` values exist for each
    variable in a data frame. Test your code on the `tibble` below.
    
```{r}
df <- tibble(
  x = c(6, NA, 9, 10, 4, 1),
  y = c("a", "z", NA, NA, NA, "e"),
  z = rnorm(6)
)
```
    
```{r}
count_na <- function(x) {
  sum(is.na(x))
}

map_dbl(df, count_na)
```

